---
title: 'World War II Flying Aces - Pilot casualties (part 3)'
author: Martin Frigaard
date: '2019-05-11'
slug: title
categories:
  - On Measurement
tags:
  - epidemiology
description: 'Casualties and aerial victories among WW2 flying aces'
topics: []
---

```{r setup, echo = FALSE, include = FALSE}
library(tidyverse)
library(skimr)
library(janitor)
# create figs
if (!file.exists("figs/")) {
    dir.create("figs/")
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " ",
  fig.path = "figs/",
  fig.height = 3,
  fig.width = 4,
  fig.align = "center",
  fig.ext = "png"
)
```


[\@mjfrigaard](https://twitter.com/mjfrigaard) here to 

> This is part 3 of a series looking at [World War II flying aces](https://en.wikipedia.org/wiki/List_of_World_War_II_flying_aces). In part 1, I showed how to download data from an a Wikipedia (html) table. Then in part 2 I showed some wrangling with `dplyr` and `janitor`, and some explortatory data analysis with `inspectdf`. We discovered that Germany had the highest average number of aerial victories.  

## Import the previous data

The data from the previous post has been loaded into Github [here](https://github.com/mjfrigaard/cloud-projects). We can use `readr` to import these data into R. 

```{r WW2FlyingAces}
WW2FlyingAces <- readr::read_csv("http://bit.ly/ww2-pilot-data")
WW2FlyingAces %>% dplyr::glimpse(78)
```

## Average number of aerial victories 

The last post ended with me showing Germany had the highest mean of WW2 flying aces. I display the top 10 mean aerial victories (`air_vic`) by `country` below.

```{r mean-aerial-victories}
WW2FlyingAces %>% 
    dplyr::group_by(country) %>% 
    dplyr::summarize(`mean aerial vicotry` = mean(air_vic)) %>% 
    dplyr::arrange(desc(`mean aerial vicotry`)) %>% 
    utils::head(10)
```

This shows the average aerial victory for `Germany` was 30 victories higher than all other countries. That's pretty wide margin. After doing some research,  



## Sign off

Thanks for reading! For more information email me at <mfrigaard@paradigmdata.io> to get in touch.

If you'd like the code that produced this blog, check out the [my GitHub repository](https://github.com/mjfrigaard/martinfrigaard-io-site).