---
title: World War II Flying Aces - Casualties by country (part 3)
author: Martin Frigaard
date: '2019-05-13'
slug: world-war-ii-flying-aces-casualties-by-country-part-3
categories:
  - R
tags:
  - wikipedia
  - tidyverse
description: ''
topics: []
---


```{r setup, echo = FALSE, include = FALSE}
library(tidyverse)
library(skimr)
library(janitor)
# create figs
# if (!file.exists("figs/")) {
#   dir.create("figs/")
# }
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = " "
  # fig.path = "figs/",
  # fig.height = 3,
  # fig.width = 5,
  # fig.align = "center",
  # fig.ext = "png"
)
```


[\@mjfrigaard](https://twitter.com/mjfrigaard) here to 

> This is part 3 of a series looking at [World War II flying aces](https://en.wikipedia.org/wiki/List_of_World_War_II_flying_aces). In part 1, I showed how to download data from an a Wikipedia (html) table. Then in part 2 I showed some wrangling with `dplyr` and `janitor`, and some exploratory data analysis with `inspectdf`. We discovered that Germany had the highest average number of aerial victories.  

## Import the previous data

The data from the previous post has been loaded into Github [here](https://github.com/mjfrigaard/cloud-projects). We can use `readr` to import these data into R. 

```{r WW2FlyingAces}
WW2FlyingAces <- readr::read_csv("http://bit.ly/ww2-pilot-data")
WW2FlyingAces %>% dplyr::glimpse(78)
```

## Average number of aerial victories 

The last post ended with me showing Germany had the highest mean of WW2 flying aces. I display the top 10 mean aerial victories (`air_vic`) by `country` below.

```{r mean-aerial-victories}
  WW2FlyingAces %>%
    dplyr::group_by(country) %>%
    dplyr::summarize(`mean aerial vicotry` = mean(air_vic)) %>%
    dplyr::mutate(`mean aerial vicotry` = round(`mean aerial vicotry`, 2)) %>% 
    dplyr::arrange(desc(`mean aerial vicotry`))
```

This shows the average aerial victory for `Germany` was 30 victories higher than all other countries. That's pretty wide margin. The average number of aerial victories might not be the best metric for determining which country had air superiority, or who had the most successful pilots. 

Tallying up the number of aerial victories might give us a picture of how good the best pilots were, but if all your best pilots are ending up in the `KIA, MIA, Died of wounds` group, you're not likely to keep winning air battles. We should also be looking at the number of casualties, or maybe even better, *the number of flying aces who survived and continued fighting throughout the entire war.*

## Flying Ace Survivability

We can use the `country` and `mia_or_kia` variables to calculate the totals and percentage of pilots who survived with the versatile `janitor::tabyl()` functions. Below are the countries with the top 10 numbers of pilots who survived. 

```{r WW2CountryXCasualties}
# create tabyl and arrange by number survived
WW2CountryXCasualties <- WW2FlyingAces %>%
  janitor::tabyl(country, mia_or_kia) %>%
  dplyr::rename(
    total_casaulty = `KIA, MIA, Died of wounds`,
    total_survived = `Lived`) %>%
  dplyr::arrange(desc(total_survived))
WW2CountryXCasualties %>% utils::head(10)
```

This shows us the `United States` and `Germany` had the highest numbers of `total_survived`. If we look over in the `total_casaulty` column, we see Germany also had a high number of pilots end up in the `KIA, MIA, Died of wounds` category. If we arrange the `WW2CountryXCasualties` `tabyl` on the `total_casaulty` column, we see that `Germany` is actually the highest ranked country in terms of total pilots `KIA, MIA, Died of wounds`.

```{r sort-total_casaulty}
WW2CountryXCasualties %>%
  dplyr::arrange(desc(total_casaulty)) %>%
  utils::head(10)
```

The totals aren't telling us enough, so we should get a ratio to compare the number of pilots who survived against those who didn't. We can do this with the `janitor::adorn_percentages("row")` function and then rename these with a more precise column and convert this back to a `tibble` for better printing. 

```{r WW2CountryXCasualtiesRatio}
# add row totals to WW2CountryXCasualties
WW2CountryXCasualtiesRatio <- WW2CountryXCasualties %>%
  janitor::adorn_percentages("row") %>%
  dplyr::rename(
    ratio_casaulty = total_casaulty,
    ratio_survived = total_survived) %>%
  as_tibble()
WW2CountryXCasualtiesRatio %>% utils::head(10)
```

Now we're getting somewhere. This is showing the percentage of pilots who survived vs. the percentage who didn't. But it would be nice to know the actual numbers of pilots in each category, so we will join these two tables back together into `WW2CountryXCasualtiesPerc`.

```{r WW2CountryXCasualtiesPerc}
# join
WW2CountryXCasualtiesPerc <- dplyr::inner_join(
  x = WW2CountryXCasualties,
  y = WW2CountryXCasualtiesRatio,
  by = "country")
WW2CountryXCasualtiesPerc %>% utils::head(10)
```

This table will be handy for plotting another Cleveland dot plot like the one from the previous post, but this time I will be able to use the ratio as a percent in the labels. 

```{r Plot-WW2CountryXCasualtiesPerc}
WW2CountryXCasualtiesPerc %>%

  # join this to the original table to get the force variable
  inner_join(x = ., y = WW2FlyingAces, by = "country") %>%

  # plot new total_survived and ratio_survived
  ggplot2::ggplot(aes(
    x = total_survived,

    # wrangle the country variable in the plot function
    y = forcats::fct_reorder(
      .f = country,
      .x = total_survived),

    # add group aesthetic here
    group = force,

    # add the scales to format the ratio_survived
    label = scales::percent(round(ratio_survived, 3)))) +

  # start this line at the country (on the y)
  ggplot2::geom_segment(aes(yend = country),

    # end at -1 on the X
    xend = -1,

    color = "grey50") +

  # add points and color by the force variable
  ggplot2::geom_point(
    size = 2.3,
    aes(color = force)) +

  # adjust size and nudge so the number is visible
  geom_text(aes(color = force),
    # adjust size
    size = 2.5,
    # nudge
    nudge_x = 32) +

  # this is for the legend on the graph
  ggplot2::scale_colour_brewer(
    palette = "Set1",
    # these are manually labelled
    limits = c("Axis", "Allied")) +

  # add this theme because it looks old-school
  ggthemes::theme_tufte(base_size = 8, ticks = 3) +

  # add labels
  ggplot2::labs(
    title = "Flying Aces Pilot Survivability in WW2",
    y = "Country/Service",
    x = "Flying aces surviving aerial victories in WW2") +

  # add theme elements
  ggplot2::theme(
    axis.text.x = element_text(
      size = 7.5,
      family = "Courier New"),
    # these look better on the axes
    axis.text.y = element_text(
      size = 7.5,
      family = "Courier New"),
    # this look better as axis titles
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    axis.title = element_text(size = 10),
    # switch the text in the legend and title
    legend.text = element_text(size = 7,
      family = "Andale Mono"),
    title = element_text(family = "Andale Mono",
      size = 12))
```

This shows the number of total number of flying aces who survived in each country, *and* the percent. This way we can get an idea for who had more air superiority throughout the war. 

Each country had their own methods and doctrine for how their pilots were trained and what counted for an aerial victory. The Japanese pilots, for example, were known for having kamikaze pilots, and built their planes with the expectation that the pilots would stay in the battle until they died. This method was moderately successful in a few battles, but also basically guaranteed that Japan would be losing pilots faster than they train and equip the replacements. Read more about this [here](https://www.latimes.com/archives/la-xpm-2004-sep-25-fg-kamikaze25-story.html).

The data above are also capable of being presented in a table using a few `janitor` functions (see below):

```{r table-WW2CountryXCasualties}
WW2CountryXCasualties %>%
  janitor::adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns()
```

This shows both numbers--those who survived and those who didn't. In the next post we will dive deeper into some specific pilots and doctrines and how they influenced the number of aerial victories.

## Sign off

Thanks for reading! For more information email me at <mfrigaard@paradigmdata.io> to get in touch.

If you'd like the code that produced this blog, check out the [my GitHub repository](https://github.com/mjfrigaard/martinfrigaard-io-site).
